---
title: "Basic Random Forest"
author: "Jie Wu"
draft: false
date: "2025-06-19"
format: 
    html:
        toc: true
        toc-depth: 2
        code-fold: false
categories: ["Machine Learning", "Data Science", "R Programming"]
---

In this post, let's run a quick random forest to find most important features that can predict WGD status. We first load the data frame from last post:

```{r}
#| message: false
#| warning: false
# Load all packages
require(tidyverse)
require(here)
require(randomForest)

# Load the mutation table from the RDS file saved in the last post
data <- readRDS(here("_data/combined_data.rds"))


```

Rather than training the random forest on all genes, we'll focus on a curated subset to improve relevance and efficiency. We'll use the MSK-IMPACT genes annoatated in OncoKB, a widely recognized resource that provides a high-confidence list of cancer-associated genes. This targeted approach streamlines computation and enhances interpretability by concentrating on genes most likely to impact our analysis.

```{r}
#| message: false
#| warning: false
file_path <- here("_data/cancerGeneList.tsv")

Gene_anno <- read_tsv(file_path, guess_max = Inf)

# pull genes in MSK-IMPACT panel 

MSK_genes <- Gene_anno %>% filter(`MSK-IMPACT` == "Yes") %>%
            pull(`Hugo Symbol`) %>% unique()

```

Now select genes in this list and make the data frame for random forest analysis

```{r}
#| message: false
#| warning: false
data_rf <- data %>% 
        dplyr::select(any_of(paste0("mut.", MSK_genes)),
                                    WGD) %>%
                filter(! is.na(WGD))

# remove "-" from names to avoid errors running RF               
names(data_rf) <- gsub("-", "", names(data_rf))

set.seed(23)
rf <- randomForest(as.factor(WGD) ~ ., 
            data_rf )

varImpPlot(rf, n.var = 20)
print(rf)

```

It's not surprising that TP53 mutation shows a strong association with WGD status in the results.

In the next post, we'll use the `caret` package to perform more standardized machine learning workflows.
